Supplementary Materials

### 2022-12-28 manuscript/supplementary.tex
more ms updates

\section{The Generalized Background Selection Models}

Under the generalized background selection models used in our likelihood, the
diversity in window $b$ is $\bar{\pi}(b, \Psi) = \bar{B}(b, \Psi) \pi_0$. Here,
$\bar{B}(b | \Psi)$ is the predicted reduction in diversity due to BGS in
window $b$, given background selection parameters $\Psi$. In practice, we
pre-calculate $B(x | \Psi)$ at fixed sites $x$ across the genome, and take the
average of the fixed sites $B$ values within widow $b$ for the average
$\bar{B}(b | \Psi)$.

GBGS models assume multiplicative fitness effects across sites. Thus, the total
reduction due to segments across the genome is the product of the individual
reductions,

@@

@@

The classic BGS model, which assumes deleterious mutations cannot fix,
expresses the reduction for a focal site in the middle of a segment
\parencite{Hudson1995-xc,Hudson1994-oh,Nordborg1996-nq}; here we consider a
focal site directly to an arbitrary side of a segment under a deleterious
mutations, and extend the model to handle a focal segment arbitrarily far from
the segment. Using our notation from equation \eqref{suppeqn:b}, the classic
mutation-selection-balance BGS model is,

@@

(c.f. \cite{Hudson1995-xc} equation 8, where the factor of two difference is
due to the position of the focal site). Often per-region rates are defined $U =
2 \mu L$ and $R = r_\text{BP} L$ are the segment-wide mutation diploid mutation
rates (mutations per diploid segment per generation) and recombination length
(Morgans).

@@

When weak deleterious mutations can fix, causing substitutions, the classic
background selection models break down. We find that the generalized background
selection model of \textcite{Santiago2016-mu} works well (see Section XXX for
our derivation). We describe how we use the models here.

\subsection{Discretization of Parameters}

In practice, to fit such a model, we must discretize both $\mu(s, k(g))$ and
$s$. Our models use a fixed $q$-element grid of selection coefficients, and
mutation rates are on a fixed $z$-element grid, which is interpolated over
during ML optimization. The number of annotation features $K$ is variable and
set by the annotation data specified. There are two ways to parameterize the
mutation rates and DFE for all annotation classes in background selection
models. First, for the \emph{free-mutation} parameterization, each annotation
feature class has a free mutation rate parameter for each of these selection
coefficients (as in equation \eqref{suppeqn:b}).

>>>>

\section{Background Selection Models}

The BGS models considered here both assume multiplicative fitness effects
across sites. Thus, the total reduction due to segments across the genome is
the product of the individual reductions,

@@

Under both background selection models used in our likelihood, the diversity in
window $b$ is $\bar{\pi}(b, \Psi) = \bar{B}(b, \Psi) \pi_0$. Here, $\bar{B}(b |
\Psi)$ is the predicted reduction in diversity due to BGS in window $b$, given
background selection parameters $\Psi$. In practice, we pre-calculate $B(x |
\Psi)$ at fixed sites $x$ across the genome, and take the average of the fixed
sites $B$ values within widow $b$ for the average $\bar{B}(b | \Psi)$.

@@

The classic BGS model, which assumes deleterious mutations cannot fix,
expresses the reduction for a focal site in the middle of a segment
\parencite{Hudson1995-xc,Hudson1994-oh,Nordborg1996-nq}; here we consider a
focal site directly to an arbitrary side of a segment under a deleterious
mutations, and extend the model to handle a focal segment arbitrarily far from
the segment. To simplify notation, we will only consider a single feature class
in this and the next section. Using our notation from equation
\eqref{suppeqn:b}, the classic mutation-selection-balance BGS model is,

@@

(c.f. \cite{Hudson1995-xc} equation 8, where the factor of two difference is
due to the position of the focal site). Often per-region rates are defined $U =
2 \mu L$ and $M = r_\text{BP} L$ are the segment-wide mutation diploid mutation
rates (mutations per diploid segment per generation) and recombination length
(Morgans).

@@

When weak deleterious mutations can fix, causing substitutions, the classic
background selection models break down. We find that the generalized background
selection model of \textcite{Santiago2016-mu} works well (see Section XXX for
our derivation). Unlike the classic BGS models which only produce B maps, that
only estimate the reduction in effective population size, The generalized BGS
model of \textcite{Santiago2016-mu} also predicts the deleterious substitution
rate per generation per segment (the ``ratchet" rate), which we call the R map.
We describe how these are calculated substitution rates per segment, for each
mutation rate and selection coefficient combination in the grids. We describe
how these are calculated here.

First, the reduction factor experienced by a focal site within the segment $g$
(and thus is similar to what selected alleles in $g$ experience) under
generalized background selection is calculated as

\begin{align}
  \widetilde{b}'(\mu, s, L, r_\text{BP}) = \frac{\widetilde{N}_{e,\infty}}{N}
\end{align}

where $\widetilde{N}_{e,\infty}$ is the asymptotic effective population size,
which is the solution to the following system of two equations (see XXX for
derivation),

\begin{align}
  {N}_{e,\infty} &= \exp \left( -\frac{V}{2} Q_\infty^2 \right) \\
  {R} &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
\end{align}

where

\begin{align}
  V &= U s - 2 {R} s \\
  Q_\infty^2 &= \frac{2}{(1-Z)(2-(2-M)Z)} \\
  Z &= 1 - \frac{Us}{U-{R}}
\end{align}

and $U = 2\mu L$ and $M = r_\text{BP} L$ as above. Solving this system of two
equations for each segment also gives the total substitution or ``ratchet" rate
per generation per segment ($\widetilde{R}$). The per-basepair substitution
rate of the segment is $\lambda = \nicefrac{\widetilde{R}}{L}$

Under Santiago and Caballero's model, the correct effective population size to
gauge reduction in heterozygosity or coalescent times is given by equation XXX.
This is because selective processes can lead to non-constant pairwise
coalescent rates, which impact pairwise coalescent rates and thus the reduction
factor. We experimented with the corrections described in Section XXX, and
found (1) they had little effect on the accuracy of the B maps, and (2) they
were incredibly costly to calculate, even with the approximations described in
equations XXX. Overall, we solve for the asymptotic $\widetilde{N}_{e,\infty}$
for each segment, and use that to calculate $b'$.

Note that the asymptotic model above determines the reduction in effective
population size experienced by a focal site in the middle of a $L$-basepair
segment, but B maps require the reduction $b'$ experienced by a focal site $x$
at an arbitrary recombination distance apart. For each segment, we pre-compute
the solution to the system of two equations, giving us $V$ and $Q_\infty^2$,
and then use these values in the equation,

\begin{align}
  b'(\mu, s, L, r, d) = \exp\left( -\frac{V}{2} \left(\frac{1}{(1-(1-d)Z}\right)^2 \right) 
\end{align}

where $Z = 1 - Us / (U-\widetilde{R})$ and $V = Us - 2\widetilde{R}s$.

\subsection{Discretization of Parameters and Annotation Feature Classes}

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g \sum_{i} b(\mu_{i, \;k(g)}, s_i, L_g, r_g, d(x, g))
\end{align}

In practice, to fit such a model, we must discretize both $\mu(s, k(g))$ and
$s$. Our models use a fixed $q$-element grid of selection coefficients, and
mutation rates are on a fixed $z$-element grid, which is later interpolated
over during the maximum likelihood optimization. The number of annotation
features $K$ is variable and set by the annotation data specified. There are
two ways to parameterize the mutation rates and DFE for all annotation classes
in background selection models. First, for the \emph{free-mutation}
parameterization, each annotation feature class has a free mutation rate
parameter for each of these selection coefficients (as in equation
\eqref{suppeqn:b}).
